Computationally Efficient Simulation of Queues: The \proglang{R} Package \pkg{queuecomputer}

### 2020-07-17 paper.tex
Check whether vectors and matricies are consistently bold.

& $\lambda$ & Rate parameter of exponential inter-arrival distribution $f_\delta = \text{exp}(\lambda)$ for $M/M/K$ queue. \\
& $\mu$ & Rate parameter of exponential service distribution $f_{\mathbf{s}} = \text{exp}(\mu)$ for $M/M/K$ queue.  \\
& $\rho := \frac{\lambda}{K \mu}$ & Traffic intensity, defined only for $M/M/K$ queues. \\ 
& $\theta_{I}$ & Parameters of arrival and service joint distribution for QDC, $(\mathbf{a}, \mathbf{s}) \sim f_{\mathbf{a}, \mathbf{s}}(\cdot | \theta_I )$. \\

@@

& $\mathbf{\delta} = (\delta_1, \ldots, \delta_n )$ & Inter-arrival process, where $\delta_i$ is $a_i - a_{i-1} \quad \forall i \in 1:n$ and $\delta_1 = a_1$. \\

@@

& $\mathbf{x} = (x_1, \ldots, x_L )$ & Change times for number of open servers. \\

& $\mathbf{y} = (y_1, \ldots, y_{L+1} )$ & Number of open servers in each epoch. \\

@@

In the case of a $M/M/K$ system equilibrium distributions for
performance measures are derived analytically, they are found in
standard queueing theory textbooks \citep{lipsky2008queueing,
  thomopoulos2012fundamentals}. For instance, the limit probability of
$N$ customers in the system $\Prob(N)$ is
%
\begin{align}
\Prob(0) &= \left[\frac{ (K \rho)^K}{K! (1 - \rho)} + 1 + \sum_{i=1}^{K-1} \frac{(K \rho)^i}{i!}   \right]^{-1}, \nonumber \\
\Prob(N) &= \begin{cases} 
\Prob(0) \frac{(K \rho)^n }{N!} & \quad  N \leq K,\\
\Prob(0) \frac{(K \rho)^n }{K! K^{N-K}} & \text{otherwise} ,
\end{cases} \label{eq:pn}
\end{align}
%
where $\rho$, the resource utilization, is defined as
$\frac{\lambda}{K \mu}$. For an $M/M/K$ system this is equal to the
expected number of busy servers divided by the total number of servers
$\frac{\E(B)}{K}$ \citep[p.~451]{cassandras2009introduction}. The
expected number of customers in the system is
\citep{bhat2015introduction}
%
\begin{align}
\E(N) = K\rho + \frac{\rho (K\rho)^K P(0)}{K!(1-\rho)^2} \label{eq:en},
\end{align}
%
and the expected waiting time is
%
\begin{align}
\E(\mathbf{w}) = \frac{(K \rho)^K P(0)}{K! K \mu(1-\rho)^2} \label{eq:ew}.
\end{align}
%
If the parameters of $f_\delta$ and $f_{\mathbf{s}}$ are uncertain,
then we must turn to predictive distributions for estimates of
performance measures, which are computed analytically for $M/M/K$
queues (Equations~\ref{eq:en} and~\ref{eq:ew}). Predictive
distributions of performance measures using Bayesian posterior
distributions are derived by \citet{armero1994bayesian,
  armero1999dealing}.

@@

In a tandem queueing network, customers traverse an ordered series of
queues before departing the system. Real examples of such systems
include airport terminals, internet services and manufacturing
systems. In a parallel network, customers are partitioned into
different $\mathbf{(a,s)}$ to be seen by separate queueing systems. In
a fork/join network a task (another term for customer) is forked into
a number of subtasks which are to be completed by distinct parallel
servers. The difference from the parallel network is that the task can
only depart the system once all subtasks have arrived at the join
point.

@@

Algorithm~\ref{alg:fixed_servers} can simulate any queue of the form $G(t)/G(t)/K/\infty/n/\mathit{FCFS}$ where $K$ and $n$ can be made arbitrarily large. Furthermore, the inter-arrival and service distributions can be of completely general form and even have a dependency structure between them. Since the arrival and service times are supplied by the user rather than sampled in-situ, the algorithm ``decouples'' statistical sampling from queue computation. This frees the user to simulate queues of arbitrarily complex $f_{\delta, \mathbf{s}}$, where $K$ is fixed.

@@

We represent the number of open servers throughout the day as a step
function. Time is on the positive real number line and is partitioned
by $L$ knot locations
$\mathbf{x} = (x_1, \ldots, x_L ) \in \mathbb{R}^L_{+}$ into $L+1$
epochs $(0, x_1], (x_1, x_2], \ldots, (x_L, \infty)$. The number of
open servers in each epoch is represented by a $L+1$ length vector
$\mathbf{y} = (y_1, \ldots, y_{L+1} ) \in \mathbb{N}^{L+1}_0$. If we
assume that none of the service times $\mathbf{s}$ span the length of
more than one epoch $(x_{l}, x_{l+1}]$, formally
%
\begin{align}
\forall i \quad \left[ s_i < \min({ x_{l+1} - x_{l} | l \in 1:L }) \right] \label{eq:condition},
\end{align}
%
then we need to consider a change in state over at most 1 knot
location. This step function is determined input by the user. Like the
arrival and service times $\mathbf{(a,s)}$ it is changeable by the
user before the simulation but not during the simulation.

@@

If Condition~\ref{eq:condition} does not hold or if, otherwise, we wish to control exactly which servers are open at what time then we must use a less computationally efficient algorithm (Algorithm~\ref{alg:changing_servers_unconditional}). Each server $k$ has its own partition of $L_k$ knot locations $\mathbf{x}_k = (x_{k,1}, \ldots, x_{k,L_k} ) \in \mathbb{R}^{L_k}_{+}$ and each $\mathbf{y}_k = (y_{k,1}, \ldots, y_{k,L_k+1} )$ is an alternating sequence of 0 and 1s of length $L_k$ indicating whether the server is open or closed respectively for the associated epoch. The vector $\mathbf{c}$ is used slightly differently to how it is used in \citet{sutton_bayesian_2011}. We use it to represent the time at which each server is next available for the current customer $i$, given the current system state $\mathbf{b}$. It is the output of the \code{next\_fun} function.

@@

This algorithm can simulate queueing systems of form
$G(t)/G(t)/K(t)/\infty/n/\mathit{FCFS}$, where $K(t)$ refers to the
number of open servers changing with time. In addition, we can specify
which particular servers are available when, not just how many and we
are not bound by Condition~\ref{eq:condition}. Once again we note that
$\mathbf{b}$ can be considered as the system state and the event list
is formed by $\mathbf{a}$ and the elements of
$\mathbf{\underline{x}}$. This function can be called with the
\code{queue\_step} function in \pkg{queuecomputer} by supplying a
`\code{server.list}' object to the \code{servers} argument. For the rest
of this paper we focus on Algorithms~\ref{alg:fixed_servers}
and~\ref{alg:changing_servers} for their relative conceptual
simplicity and computational efficiency.

@@

In \pkg{simmer} and \pkg{simpy} users supply generator functions for
simulating $\mathbf{\delta}$ and service times $\mathbf{s}$, the user
enters the set of input parameters $\theta_I$ for these generator
functions and starts the simulation. The inter-arrival time is
re-sampled after each arrival and the service time is sampled when the
server begins with a new customer. This makes it difficult to model
queues where distributions for inter-arrival times do not make sense:
like the immigration counter for an airport, where multiple flights
generate customers; or when arrival times and service times are not
independent. In \pkg{queuecomputer} sampling is ``decoupled'' from
computation, the user samples $\mathbf{a}$ and service times
$\mathbf{s}$ using any method. The outputs $\mathbf{d}$ and
$\mathbf{p}$ are then computed deterministically.

@@

To demonstrate the validity of the algorithm we consider a
$M/M/2/\infty/1000/\mathit{FCFS}$ queue. If QDC is valid for any
$M/M/K$ queueing system, then it is valid for any $G(t)/G(t)/K$
queueing system. This is because any non-zero $\mathbf{(a,s)}$ could
conceivably come from two exponential distributions, even if the
probability of the particular realization is vanishingly small. We
replicate exact departure times computed with the \pkg{simmer} and
\pkg{simpy} packages using \pkg{queuecomputer}. First, we generate
$\mathbf{a}$ and $\mathbf{s}$ to be used as input to all three
packages.
%
\begin{CodeChunk}
\begin{CodeInput}
R> set.seed(1)
R> n_customers <- 10^4
R> lambda_a <- 1/1
R> lambda_s <- 1/0.9
R> interarrivals <- rexp(n_customers, lambda_a)
R> arrivals <- cumsum(interarrivals)
R> service <- rexp(n_customers, lambda_s)
\end{CodeInput}  
\end{CodeChunk}
%
We now input these objects into the three scripts using \pkg{queuecomputer}, \pkg{simmer}, or \pkg{simpy}. First, we run the \pkg{queuecomputer} script. The \code{queuecomputer_output} object is sorted in ascending order so that the departure times can be compared to the DES packages.
%
\begin{CodeChunk}
\begin{Sinput}
R> queuecomputer_output <- queue_step(arrivals = arrivals, 
+    service = service, servers = 2)
R> head(sort(depart(queuecomputer_output)))
\end{Sinput}
\begin{Soutput}
[1] 1.340151 2.288112 2.639976 2.796572 3.249794 5.714967
\end{Soutput}
\end{CodeChunk}
%
The DES packages \pkg{simmer} and \pkg{simpy} are not built to allow
users to input $(\mathbf{a,s})$ directly. Rather, the user supplies
parameters for $f_{\delta}$ and $f_{\mathbf{s}}$ so that inter-arrival
and service times can be sampled at each step when needed. To allow
\pkg{simmer} and \pkg{simpy} to accept pre-sampled input
$(\mathbf{a,s})$ we use generator functions instead of
\code{rexp(rate)} or \code{random.expovariate(rate)} calls in
\proglang{R} and \proglang{Python} respectively, details of this work
can be found in the supplementary material. We create an interface to
\pkg{simmer} so that it can be called in the same way as
\pkg{queuecomputer}.
%
\begin{CodeChunk}
\begin{Sinput}
R> simmer_output <- simmer_step(arrivals = arrivals, 
+    service = service, servers = 2)
R> head(simmer_output)
\end{Sinput}
\begin{Soutput}
[1] 1.340151 2.288112 2.639976 2.796572 3.249794 5.714967
\end{Soutput}
\end{CodeChunk}
%
The same departure times are observed. Similarly in \proglang{Python}
we create an interface to \pkg{simpy} so that it can be called in a
similar way to \pkg{queuecomputer}.
%
\begin{CodeChunk}
\begin{CodeInput}
>>> simpy_step(interarrivals, service)[0:6]
\end{CodeInput}
\begin{CodeOutput}
array([ 1.34015149,  2.28811237,  2.63997568,  2.79657232,  3.24979406, 
	5.7149671 ])
\end{CodeOutput}
\end{CodeChunk}
%
A check of all three sorted vectors of $\mathbf{d}$ from each package
revealed that all were equal to within 5 significant figures for every
$d_i, i = 1:1000$.

>>>>

& $\lambda$ & Rate parameter of exponential inter-arrival distribution $f_{\boldsymbol{\delta}} = \text{exp}(\lambda)$ for $M/M/K$ queue. \\
& $\mu$ & Rate parameter of exponential service distribution $f_{\mathbf{s}} = \text{exp}(\mu)$ for $M/M/K$ queue.  \\
& $\rho := \frac{\lambda}{K \mu}$ & Traffic intensity, defined only for $M/M/K$ queues. \\ 
& $\theta_{I}$ & Parameters of arrival and service joint distribution for QDC, $(\mathbf{a}, \mathbf{s}) \sim f_{\mathbf{a}, \mathbf{s}}(\cdot | \theta_I )$. \\

@@

& $\boldsymbol{\delta} = (\delta_1, \ldots, \delta_n )$ & Inter-arrival process, where $\delta_i$ is $a_i - a_{i-1} \quad \forall i \in 1:n$ and $\delta_1 = a_1$. \\

@@

& $\mathbf{x} = (x_1, \ldots, x_L )$ & Change times for number of open servers. Input for Algorithm \ref{alg:changing_servers}. \\

& $\mathbf{y} = (y_1, \ldots, y_{L+1} )$ & Number of open servers in each epoch. Input for Algorithm \ref{alg:changing_servers}.  \\

& $\mathbf{\underline{x}} = (\mathbf{x}_1, \ldots, \mathbf{x}_K )$ & List of change times for each server. Input for Algorithm \ref{alg:changing_servers_unconditional}. \\

& $\mathbf{\underline{y}} = (\mathbf{y}_1, \ldots, \mathbf{y}_K )$ & List with a Boolean vector for each server. A zero indicates that the server is open, a one indicates that the server is closed for the corresponding epoch in $\mathbf{\underline{x}}$. Input for Algorithm \ref{alg:changing_servers_unconditional}. \\

@@

In the case of a $M/M/K$ system equilibrium distributions for
performance measures can be derived analytically. These derivations are 
found in standard queueing theory textbooks \citep{lipsky2008queueing,
  thomopoulos2012fundamentals}. For instance, the limit probability of
$N$ customers in the system $\Prob(N)$ is
%
\begin{align}
\Prob(0) &= \left[\frac{ (K \rho)^K}{K! (1 - \rho)} + 1 + \sum_{i=1}^{K-1} \frac{(K \rho)^i}{i!}   \right]^{-1}, \nonumber \\
\Prob(N) &= \begin{cases} 
\Prob(0) \frac{(K \rho)^n }{N!} & \quad  N \leq K,\\
\Prob(0) \frac{(K \rho)^n }{K! K^{N-K}} & \text{otherwise} ,
\end{cases} \label{eq:pn}
\end{align}
%
where $\rho$, the resource utilization, is defined as
$\frac{\lambda}{K \mu}$. For an $M/M/K$ system this is equal to the
expected number of busy servers divided by the total number of servers
$\frac{\E(B)}{K}$ \citep[p.~451]{cassandras2009introduction}. The
expected number of customers in the system is
\citep{bhat2015introduction}
%
\begin{align}
\E(N) = K\rho + \frac{\rho (K\rho)^K P(0)}{K!(1-\rho)^2} \label{eq:en},
\end{align}
%
and the expected waiting time is
%
\begin{align}
\E(\mathbf{w}) = \frac{(K \rho)^K P(0)}{K! K \mu(1-\rho)^2} \label{eq:ew}.
\end{align}
%
If the parameters of $f_{\boldsymbol{\delta}}$ and $f_{\mathbf{s}}$ are uncertain,
then we must turn to predictive distributions for estimates of
performance measures, which are computed analytically for $M/M/K$
queues (Equations~\ref{eq:en} and~\ref{eq:ew}). Predictive
distributions of performance measures using Bayesian posterior
distributions are derived by \citet{armero1994bayesian,
  armero1999dealing}.

@@

In a tandem queueing network, customers traverse an ordered series of
queues before departing the system. Real examples of such systems
include airport terminals, internet services and manufacturing
systems. In a parallel network, customers are partitioned into
different queueing systems, with their own two vectors $\mathbf{a}$ and $\mathbf{s}$.
In a fork/join network a task (another term for customer) is split into
a number of subtasks which are to be completed by distinct parallel
servers. The task is only complete once all its associated subtasks arrive
at the join point. One example of this situation is that of arriving passengers
and their bags at an airport.

@@

Algorithm~\ref{alg:fixed_servers} can simulate any queue of the form $G(t)/G(t)/K/\infty/n/\mathit{FCFS}$ where $K$ and $n$ can be made arbitrarily large. Furthermore, the inter-arrival and service distributions can be of completely general form and even have a dependency structure between them. Since the arrival and service times are supplied by the user rather than sampled in-situ, the algorithm ``decouples'' statistical sampling from queue computation. This frees the user to simulate queues of arbitrarily complex $f_{\boldsymbol{\delta}, \mathbf{s}}$, where $K$ is fixed.

@@

We represent the number of open servers throughout the day as a step
function. Time is on the positive real number line and is partitioned
by $L$ knot locations
$\mathbf{x} = (x_1, \ldots, x_L ) \in \mathbb{R}^L_{+}$ into $L+1$
epochs $(0, x_1], (x_1, x_2], \ldots, (x_L, \infty)$. The number of
open servers in each epoch is represented by a $L+1$ length vector
$\mathbf{y} = (y_1, \ldots, y_{L+1} ) \in \mathbb{N}^{L+1}_0$. If we
assume that none of the service times $\mathbf{s}$ span the length of
more than one epoch $(x_{l}, x_{l+1}]$, formally
%
\begin{align}
\forall i \quad \left[ s_i < \min({ x_{l+1} - x_{l} | l \in 1:L }) \right] \label{eq:condition},
\end{align}
%
then we need to consider a change in state over at most 1 knot
location. This step function is determined input by the user. Like the
arrival and service times $\mathbf{(a,s)}$, it is changeable by the
user before the simulation but not during the simulation.

@@

If Condition~\ref{eq:condition} does not hold or if, otherwise, we wish to control exactly which servers are open at what time then we must use a less computationally efficient algorithm (Algorithm~\ref{alg:changing_servers_unconditional}). Each server $k$ has its own partition of $L_k$ knot locations $\mathbf{x}_k = (x_{k,1}, \ldots, x_{k,L_k} ) \in \mathbb{R}^{L_k}_{+}$ and each $\mathbf{y}_k = (y_{k,1}, \ldots, y_{k,L_{k+1}} )$ is an alternating sequence of 0 and 1s of length $L_{k+1}$ indicating whether the server is open or closed respectively for the associated epoch. The vector $\mathbf{c}$ is used slightly differently to how it is used in \citet{sutton_bayesian_2011}. We use it to represent the time at which each server is next available for the current customer $i$, given the current system state $\mathbf{b}$. It is the output of the \code{next\_fun} function.

@@

This algorithm can simulate queueing systems of form
$G(t)/G(t)/K(t)/\infty/n/\mathit{FCFS}$, where $K(t)$ refers to the
number of open servers changing with time. In addition, we can specify
which particular servers are available when, not just how many and we
are not bound by Condition~\ref{eq:condition}. Once again we note that
$\mathbf{b}$ can be considered as the system state and the event list
is formed by $\mathbf{a}$ and $\mathbf{x}_1, \dots, \mathbf{x}_K$, where 
$K$ in this context is the total number of distinct servers.
This function can be called with the
\code{queue\_step} function in \pkg{queuecomputer} by supplying a
`\code{server.list}' object to the \code{servers} argument. For the rest
of this paper we focus on Algorithms~\ref{alg:fixed_servers}
and~\ref{alg:changing_servers} for their relative conceptual
simplicity and computational efficiency.

@@

In \pkg{simmer} and \pkg{simpy} users supply generator functions for
simulating $\boldsymbol{\delta}$ and service times $\mathbf{s}$, the user
enters the set of input parameters $\theta_I$ for these generator
functions and starts the simulation. The inter-arrival time is
re-sampled after each arrival and the service time is sampled when the
server begins with a new customer. This makes it difficult to model
queues where distributions for inter-arrival times do not make sense:
like the immigration counter for an airport, where multiple flights
generate customers; or when arrival times and service times are not
independent. In \pkg{queuecomputer} sampling is ``decoupled'' from
computation, the user samples $\mathbf{a}$ and service times
$\mathbf{s}$ using any method. The outputs $\mathbf{d}$ and
$\mathbf{p}$ are then computed deterministically.

@@

To demonstrate the validity of the algorithm we consider a
$M/M/2/\infty/1000/\mathit{FCFS}$ queue. If QDC is valid for any
$M/M/K$ queueing system, then it is valid for any $G(t)/G(t)/K$
queueing system. This is because any non-zero $\mathbf{(a,s)}$ could
conceivably come from two exponential distributions, even if the
probability of the particular realization is vanishingly small. We
replicate exact departure times computed with the \pkg{simmer} and
\pkg{simpy} packages using \pkg{queuecomputer}. First, we generate
$\mathbf{a}$ and $\mathbf{s}$ to be used as input to all three
packages.
%
\begin{CodeChunk}
\begin{CodeInput}
R> set.seed(1)
R> n_customers <- 10^4
R> lambda_a <- 1/1
R> lambda_s <- 1/0.9
R> interarrivals <- rexp(n_customers, lambda_a)
R> arrivals <- cumsum(interarrivals)
R> service <- rexp(n_customers, lambda_s)
\end{CodeInput}  
\end{CodeChunk}
%
We now input these objects into the three scripts using \pkg{queuecomputer}, \pkg{simmer}, or \pkg{simpy}. First, we run the \pkg{queuecomputer} script. The \code{queuecomputer_output} object is sorted in ascending order so that the departure times can be compared to the DES packages.
%
\begin{CodeChunk}
\begin{Sinput}
R> queuecomputer_output <- queue_step(arrivals = arrivals, 
+    service = service, servers = 2)
R> head(sort(depart(queuecomputer_output)))
\end{Sinput}
\begin{Soutput}
[1] 1.340151 2.288112 2.639976 2.796572 3.249794 5.714967
\end{Soutput}
\end{CodeChunk}
%
The DES packages \pkg{simmer} and \pkg{simpy} are not built to allow
users to input $(\mathbf{a,s})$ directly. Rather, the user supplies
parameters for $f_{\boldsymbol{\delta}}$ and $f_{\mathbf{s}}$ so that inter-arrival
and service times can be sampled at each step when needed. To allow
\pkg{simmer} and \pkg{simpy} to accept pre-sampled input
$(\mathbf{a,s})$ we use generator functions instead of
\code{rexp(rate)} or \code{random.expovariate(rate)} calls in
\proglang{R} and \proglang{Python} respectively, details of this work
can be found in the supplementary material. We create an interface to
\pkg{simmer} so that it can be called in the same way as
\pkg{queuecomputer}.
%
\begin{CodeChunk}
\begin{Sinput}
R> simmer_output <- simmer_step(arrivals = arrivals, 
+    service = service, servers = 2)
R> head(simmer_output)
\end{Sinput}
\begin{Soutput}
[1] 1.340151 2.288112 2.639976 2.796572 3.249794 5.714967
\end{Soutput}
\end{CodeChunk}
%
The same departure times are observed. Similarly in \proglang{Python}
we create an interface to \pkg{simpy} so that it can be called in a
similar way to \pkg{queuecomputer}.
%
\begin{CodeChunk}
\begin{CodeInput}
>>> simpy_step(interarrivals, service)[0:6]
\end{CodeInput}
\begin{CodeOutput}
array([ 1.34015149,  2.28811237,  2.63997568,  2.79657232,  3.24979406, 
	5.7149671 ])
\end{CodeOutput}
\end{CodeChunk}
%
A check of all three sorted vectors of $\mathbf{d}$ from each package
revealed that all were equal to within 5 significant figures for every
$d_i, i = 1:1000$.
